Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-13 programlogic.tex
+

\todo[inline]{Cite Kolanski here; not all of them, but some actually defined it exactly this way, for the current
address space.}
\subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} 
  is too strong to specify some operations a virtual memory manager may need to do, such as
  move one level of the page table to a different physical location while preserving all virtual-to-physical mappings.\footnote{
    x86-64 hardware, like other architectures, includes a feature (which we do not formalize assertions for) to
    replace an L1 page table address in an L2 entry with a pointer to a \emph{larger} 2MB page (called super-pages), 
    or replace an L2 page table address in an L3 entry with a pointer to a 1GB page (called huge-pages).
  }
   To speak more concretely, the fractional permissions on the page-table-walk assertion ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}) 
  ensure fractional ownership of the four physical page table entries accessed by the hardware during address translation,
  as well as the physical memory holding the data of the virtual points-to.
  However, this formulation stores knowledge of the page table walk's details with ownership of the backing physical
  memory.
  Updating any of these mappings (e.g., move the page tables in physical memory, as in coallescing for
  superpages or hugepages)
  would require explicitly collecting all virtual points-to facts that traverse affected entries.
  It is preferable to permit the page tables themselves to be updated independently of the virtual points-to assertions,
  so long as those updates preserve the same virtual-to-physical translations.

\begin{figure*}
\todo[inline]{Colin says: again, this is supposed to be a vprop, but isn't doing anything with a cr3val argument.
I realize the GASI and per-address-space invariants, and therefore machinery to tie the virtual points-to
to the GASI, hasn't been introduced yet. Is there a short-hand here we can use to be ``more complete'' until the later text
that fills in these details?
}
\todo[inline,color=cyan]{Colin says: Oh!!! I see. You have that extremely subtle rtv subscript on the virtual points-to
syntax! Don't do that. First, it's really subtle and wasn't explained anywhere, so even I missed it. Second,
we're very explicitly describing vProps as functions from root to iProp. Sticking the rtv on the arrow there
is partly weird because it obscures which argument is really the vProp argument (there are 2 other word64s in that syntax),
partly weird because it's written next to something that \emph{isn't} a variable (the v subscript on the virtual points-to
arrow) while using the same font, and partly weird because when you actually write these assertions, rtv doesn't show
up in the syntax at all, which is the whole point --- virtual points-to defs shouldn't mention their cr3 or address space at
all!
Here's what I think it should look like (I also existentially quantified $\delta$ and fixed the p subscript on physical
points-to):
}
\[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{v}}\{\textsf{q}\}\vpage : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \lambda \mathit{cr3val}\ldotp
    \exists \paddr\ldotp
    \exists \delta\ldotp
    \sumapaces{\mathit{cr3val}}\delta \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vpage
\end{array}
\]
\todo[inline]{Colin says: The fact that we can't visually distinguish the variable v that is the value in memory
from the subscript v on the virtual points-to arrow is not ideal: it makes it hard to tell what's concrete syntax
and what's a variable. This is worth fixing, but near the bottom of the priority list}
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}

@@

  \begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{l1e})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\textsf{ maddr, l4e, l3e, l2e, l1e})} \\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammPred \urcorner \ast
    \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}

>>>>

\todo[inline]{Cite Kolanski here; not all of them, but some actually defined it exactly this way, for the current
address space.}
\subsection{Aliasing/Sharing Physical Pages}
  \label{sec:sharingpages}  
  One might have already observed that the virtual points-to definition shown in Figure \ref{fig:strongvirtualpointsto} 
  is too strong to specify some operations a virtual memory manager may need to do, such as
  move one level of the page table to a different physical location while preserving all virtual-to-physical mappings.\footnote{
    x86-64 hardware, like other architectures, includes a feature (which we do not formalize assertions for) to
    replace an L1 page table address in an L2 entry with a pointer to a \emph{larger} 2MB page (called super-pages), 
    or replace an L2 page table address in an L3 entry with a pointer to a 1GB page (called huge-pages).
  }
   To speak more concretely, the fractional permissions on the page-table-walk assertion ($\textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}$ in Figure \ref{fig:strongvirtualpointsto}) 
  ensure fractional ownership of the four physical page table entries accessed by the hardware during address translation,
  as well as the physical memory holding the data of the virtual points-to.
  However, this formulation stores knowledge of the page table walk's details with ownership of the backing physical
  memory.
  Updating any of these mappings (e.g., move the page tables in physical memory, as in coallescing for
  superpages or hugepages)
  would require explicitly collecting all virtual points-to facts that traverse affected entries.
  It is preferable to permit the page tables themselves to be updated independently of the virtual points-to assertions,
  so long as those updates preserve the same virtual-to-physical translations.
\newcommand{\vale}{\textsf{val}}
\begin{figure*}
\todo[inline]{Colin says: again, this is supposed to be a vprop, but isn't doing anything with a cr3val argument.
I realize the GASI and per-address-space invariants, and therefore machinery to tie the virtual points-to
to the GASI, hasn't been introduced yet. Is there a short-hand here we can use to be ``more complete'' until the later text
that fills in these details?
}
\todo[inline,color=cyan]{Colin says: Oh!!! I see. You have that extremely subtle rtv subscript on the virtual points-to
syntax! Don't do that. First, it's really subtle and wasn't explained anywhere, so even I missed it. Second,
we're very explicitly describing vProps as functions from root to iProp. Sticking the rtv on the arrow there
is partly weird because it obscures which argument is really the vProp argument (there are 2 other word64s in that syntax),
partly weird because it's written next to something that \emph{isn't} a variable (the v subscript on the virtual points-to
arrow) while using the same font, and partly weird because when you actually write these assertions, rtv doesn't show
up in the syntax at all, which is the whole point --- virtual points-to defs shouldn't mention their cr3 or address space at
all!
Here's what I think it should look like (I also existentially quantified $\delta$ and fixed the p subscript on physical
points-to):
}
\[
\begin{array}{l}
    \vaddr\mapsto_{\textsf{v}}\{\textsf{q}\}\vale : \mathsf{vProp}~\Sigma \stackrel{\triangle}{=} 
    \exists \paddr\ldotp
    \exists \delta\ldotp
    \sumapaces{\mathit{cr3val}}\delta \ast 
  \sumwalkabs\vaddr\qfrac\paddr \ast \paddr \mapsto_{\mathsf{p}} \vale
\end{array}
\]
\todo[inline]{Colin says: The fact that we can't visually distinguish the variable v that is the value in memory
from the subscript v on the virtual points-to arrow is not ideal: it makes it hard to tell what's concrete syntax
and what's a variable. This is worth fixing, but near the bottom of the priority list}
\caption{Virtual-Pointsto for Sharing Pages}
  \label{fig:virtualpointstosharing}
\end{figure*}

@@

  \begin{figure*}
\[
\begin{array}{l}
  \mathcal{I}\textsf{ASpace}(\ptablestore,m)\stackrel{\triangle}{=} \textsf{ASpace\_Lookup}(\ptablestore,m) \ast \\
 \qquad\qquad\qquad \bigast{(\vaddr, \textsf{l1e})\in \ptablestore}{\exists\;(\textsf{l4e l3e l2e})\ldotp \textsf{L}_{4}\_\textsf{L}_{1}\_\textsf{PointsTo}(\textsf{ maddr, l4e, l3e, l2e, l1e})} \\
  \textsf{ where } \\
   \textsf{ASpace\_Lookup} (\ptablestore,m) \stackrel{\triangle}{=} \lambda\textsf{ cr3val} \ldotp \; \ulcorner m \; !!\; \textsf{cr3val} = \textsf{Some } \gammaPred \urcorner \ast
    \ownGhost\gammaPred{\authfull{\ptableabswalk\ptablestore}}
